Tidy data structure to support exploration and modeling of temporal data

### 2018-12-09 main.Rmd
restarted tidy data

Suppose that temporal data consists of $N$ subjects/observational units. Given subject $i$ $(i = 1, 2, \dots, N)$, $X_{ijt}$ denotes the $j$th measurement $(j = 1, 2, \dots, p)$ indexed at a unique sequence of time points $1 \le t \le T_{i}$. Temporal data arrives in many possible formats, so does time. For example, time can carry with various time resolutions (hours, minutes, and seconds), and they are typically associated with different time zones with adjustments like summer time. It could be irregularly recorded, which is particularly true with longitudinal measurements like patient visits to a doctor's office. Temporal data also often contains rich information: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors. An overview of data science workflow is given in Figure \ref{fig:pipeline-ann}. The divide between data that feeds into the beginning of analysis and the current time series standard that fits into modelling is significant, making data preparation for modeling and forecasting painful.

@@

@@

@wickham2014tidy coined the term "tidy data", which is a rephrasing of the second and third normal forms in relational databases but in a way that makes more sense to data scientists. A suite of paired verbs such as `gather` and `spread` [@R-tidyr] were used to describe the reshape processing from messy data to tidy data. Tidy data, which standardized the mapping from the semantics of a dataset to its physical representation, serves remarkably as a fundamental unit of data analysis. R package **ggplot2** [@wickham2009ggplot2] pioneered developing the grammar of graphics based on tidy data, thus making it possible mapping from data space to visual elements. **dplyr** [@R-dplyr] generalised a coherent and consistent set of verbs to handle a wide range of data transformation tasks. @r4ds argues that 80% of data analysis tasks can be solved with tidy tools while the remaining 20% requires other tools. [R] The prevalence of grammar

@@

\begin{table}[!hbt]
\caption{A test}\label{tab:verb-table}
\centering
\begin{tabular}{ lll }
\toprule
 & Verb & Description \\ \midrule
\multirow{3}{*}{Time gaps} & \texttt{has\_gaps()} & Test if a tsibble has gaps in time \\
 & \texttt{count\_gaps()} & Count and report time gaps \\
 & \texttt{fill\_gaps()} & Fill in gaps by values and functions \\ \hline
\multirow{4}{*}{Row-wise} & \textbf{\texttt{filter()}} & Pick rows based on conditions \\
 & \texttt{filter\_index()} & Provide a shorthand for time subsetting \\
 & \textbf{\texttt{slice()}} & Select rows based on row positions \\
 & \textbf{\texttt{arrange()}} & Sort the ordering of row by variables \\ \hline
\multirow{4}{*}{Column-wise} & \textbf{\texttt{select()}} & Pick columns by variables \\
 & \textbf{\texttt{mutate()}} & Add new variables \\
 & \textbf{\texttt{transmute()}} & Add new variables \\
 & \textbf{\texttt{summarise()}} & Aggregate values over time \\ \hline
\multirow{3}{*}{Group-wise} & \texttt{index\_by()} & Group by new time index \\
 & \textbf{\texttt{group\_by()}} & Group by one or more variables \\
 & \texttt{group\_by\_key()} & Group by key variables \\ \hline
\multirow{4}{*}{Reshape data} & \textbf{\texttt{gather()}} & David Batty \\
 & \textbf{\texttt{spread()}} & Eirik Bakke \\
 & \textbf{\texttt{nest()}} & Eirik Bakke \\
 & \textbf{\texttt{unnest()}} & Jody Morris \\
\bottomrule
\end{tabular}
\end{table}

@@

>>>>

Suppose that temporal data consists of $N$ subjects/observational units. Given subject $i$ $(i = 1, 2, \dots, N)$, $X_{ijt}$ denotes the $j$th measurement $(j = 1, 2, \dots, p)$ indexed at a unique sequence of time points $0 \le t \le T_{i}$. Temporal data arrives in many possible formats, so does time. For example, time can carry with various time resolutions (hours, minutes, and seconds), and they are typically associated with different time zones with adjustments like summer time. It could be irregularly recorded, which is particularly true with longitudinal measurements like patient visits to a doctor's office. Temporal data also often contains rich information: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors. @r4ds succinctly describes the data science workflow given in Figure \ref{fig:pipeline-ann}. It is annotated to show that the divide between data that feeds into the beginning of analysis and the current time series objects that fits into modelling is significant, making data preparation for modeling and forecasting painful.

@@

The existing standards, particularly time series objects, are model-oriented with little support for non-numeric values, leaning its data abstraction towards matrix. Time series have been isolated from general data problems due to this data choice. Analysts who start analysing time series would be pushed to learn new transformation and visualization toolkits that are different from what they used to operate.

@@

@wickham2014tidy coined the term "tidy data", which is a rephrasing of the second and third normal forms in relational databases but in a way that makes more sense to data scientists by referring rows to observations and columns to variables. The principles of "tidy data" attempt to standardize the mapping of the semantics of a dataset to its physical representation. This data structure is the fundamental unit of the **tidyverse**, which is a collection of R packages designed for data science; the prevalence of **tidyverse** reversely testifies to the novelty of "tidy data". The **tidyverse** provides abstract yet functional grammars to manipulate and visualize data in easier-to-comprehend manners. One of the **tidyverse** packages, **dplyr** [@R-dplyr] showcases the value of a grammar as a principled vehicle to transform data for a wide range of data challenges, providing a consistent set of verbs: `mutate()`, `select()`, `filter()`, `summarise()`, and `arrange()`. Each verb focuses on a singular task. Almost all data tasks can be rephrased and tackled with these five key verbs, by composing them sequentially.

Community effort for building tidy ecosystem

Temporal data is integrated to tidy data and the grammar. @r4ds argues that 80% of data analysis tasks can be solved with tidy tools while the remaining 20% requires other tools.

@@

\begin{table}[!hbt]
\caption{A test}\label{tab:verb-table}
\centering
\begin{tabular}{ lll }
\toprule
 & Verb & Description \\ \midrule
\multirow{4}{*}{Time gaps} & \texttt{has\_gaps()} & Test if a tsibble has gaps in time \\
 & \texttt{scan\_gaps()} & Report implicit missing entries \\
 & \texttt{count\_gaps()} & Summarize time gaps \\
 & \texttt{fill\_gaps()} & Fill in gaps by values and functions \\ \hline
\multirow{4}{*}{Row-wise} & \textbf{\texttt{filter()}} & Pick rows based on conditions \\
 & \texttt{filter\_index()} & Provide a shorthand for time subsetting \\
 & \textbf{\texttt{slice()}} & Select rows based on row positions \\
 & \textbf{\texttt{arrange()}} & Sort the ordering of row by variables \\ \hline
\multirow{4}{*}{Column-wise} & \textbf{\texttt{select()}} & Pick columns by variables \\
 & \textbf{\texttt{mutate()}} & Add new variables \\
 & \textbf{\texttt{transmute()}} & Add new variables \\
 & \textbf{\texttt{summarise()}} & Aggregate values over time \\ \hline
\multirow{3}{*}{Group-wise} & \texttt{index\_by()} & Group by new time index \\
 & \textbf{\texttt{group\_by()}} & Group by one or more variables \\
 & \texttt{group\_by\_key()} & Group by key variables \\ \hline
\multirow{4}{*}{Reshape data} & \textbf{\texttt{gather()}} & David Batty \\
 & \textbf{\texttt{spread()}} & Eirik Bakke \\
 & \textbf{\texttt{nest()}} & Eirik Bakke \\
 & \textbf{\texttt{unnest()}} & Jody Morris \\
\bottomrule
\end{tabular}
\end{table}

@@

```{r elec-gaps}
```
